Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-08 pkg/disttree/inst/draft/draft.Rnw
revised 2nd part of intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1989 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4e1723563d4c1e5ffbb1f3395fabdef2a3826325

In regression analysis various models have been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1986})
    encompassing an additional nonlinear link function for the conditional mean.
  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item Generalized additive models for location, scale, and shape
    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

For these reasons, regression trees and particularly random forests provide
valuable features which can contribute to improving models for probabilistic
forecasting. Therefore, we are going to introduce a novel general distributional
framework for these methods by presenting distributional trees and forests which
embed the idea of flexible parametric modeling into the framework of regression
trees and random forests. In Figure~\ref{Devel_treeforest} the developments in
the field of tree-structured models together with the novel approaches of
distributional trees and forests are illustrated for the case of one covariate.
In the top left picture the step function modeling the mean of the response
variable represents the division into three terminal subgroups by a regression
tree. This is extended in the bottom left picture where a distributional tree
models not only the mean but also the variance of the response variable by its
own step function. The top middle picture shows that combining an ensemble of
trees to a random forest can lead to smoother effects while the tree structure
assures that abrupt changes are captured as well. Again, an extension of the
random forest model considering only the mean is made in the top right picture.
Here a distributional forest models the mean and the variance capturing abrupt
changes but also smooth effects.

However, for example in complex
terrain in mountainous regions various factors can change rapidly and their
effects are difficult to be represented by a smooth parametric model. In this
situation a different approach which can deal with these abrupt changes would be
to apply tree-structured models such as regression trees
(\citealt{breiman1984classification}) or random forests such as Breiman and
Cutler's randomForest (\citealt{breiman2001random}). Moreover, these models are
particularly advantageous in case of many covariates with unknown effects or
interactions as the variable selection is done automatically within the model
fitting process. Therefore, no additional effort or knowledge for the
specification of the model is required in advance. However, even though their
main advantage of being able to model non-additive effects and interactions
without any necessary prespecification can play to their strength, regression
trees are not an adequate choice if (additional) smooth effects appear as well.
In this situation a random forest provides a good compromise as it combines a
set of trees such that the steps imposed by the trees are smoothed out in case
of smooth effects while abrupt changes are still captured by the tree structure
of the model.

@@

After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}. The new method is applied on precipitation data in a mountainous region with a very large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area. In a first setting only one observation station is taken into consideration. The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a distributional forest are plotted together with the corresponding observations. This illustrates that well fitting predicted expected values 
%%% well fitting predicted expected values -> appropriate probabilistic forecasts? </TH>
are provided by this model as they are close to the observed values. Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis. In section \ref{ProbabilisticForecasting} this example will be further discussed and it will be shown that the predictions made by the distributional forest are reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, the scale parameter, and therefore the whole distribution.

>>>>

In regression analysis various models have been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1983})
    encompassing an additional nonlinear link function for the conditional mean.
  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item Generalized additive models for location, scale, and shape
    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

The before discussed models with the response depending on additive linear or smooth nonlinear terms
of the covariates are powerful tools as long as only smooth effects are to be considered.
A different approach that can also deal with non-additive effects
and possibly unknown interactions of covariates is to apply tree-based methods. 
A regression tree (\citealt{breiman1984classification})
works on the idea of separating the data into more homogeneous subgroups based on
covariates. However, the resulting step function as illustrated in the left panel of
Figure \ref{Devel_treeforest} can often be too rough and imposes abrupt changes on the model/data even in
case of smooth transitions. Random forest models such as Breiman and
Cutler's randomForest (\citealt{breiman2001random}) encounter this problem/disadvantage by combining an
ensemble of trees such that steps are smoothed out in case of smooth effects while abrupt changes
are still captured by its tree-based structure (see the top middel panel of Figure \ref{Devel_treeforest}).
While classic regression trees and random forests only model the mean of the response
we are now going to combine the idea of these tree based-methods with the progress
made in the field of (flexible) parametric modelling. In particular, the development from mean regression
models to distributional models will be embeded in the framework or regression trees and
random forests providing distributional trees and forests. This development is illustrated
in the bottom panel of Figure~\ref{Devel_treeforest} for distributional trees and in the top right panel
for distributional forests.

In many modeling situations, especially when the type of effects are not known in advance,
it can be challenging to specify a well fitting model. The novel methodology of distributional 
trees and forests provides a framework that covers a wide range of effects including linear and nonlinear
additive and non-additive effects as well as interactions of covariates. Moreover, these models are
particularly advantageous in case of many covariates with unknown effects or
interactions as the variable selection is done automatically within the model
fitting process. Therefore, no additional effort or knowledge for the
specification of the model is required in advance.

For these reasons, distributional trees and forests provide
valuable features which can contribute to improving models for probabilistic
forecasting. For example in precipitation/weather forecasting (for complex terrain) 
in mountainous regions various influencing factors can change rapidly and their effects are 
difficult to be represented by a smooth parametric model. (Additionally, a wide grid over
this complex terrain can lead to abrupt changes within just one cell.)

In the application discussed in this paper precipitation forecasts are made for a mountainous region
in the Alps covering mainly Tyrol, Austria, and adjacent areas. Predictions are based on data from 
85 observation stations and 28 years (1985--2012). The total amount of precipitation within 24 hours
is modeled with the assumption of the underlying distribution being a censored normal distribution 
with censoring point 0. As covariates a large number of meteorlogical quantities are provided by 
ensembles of numerical weather prediction models. 
Next to distributional other methods commonly used in this research area are applied and their results are compared. 
In a first setting only one observation station is taken into consideration. The models are trained on data from 
24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of 
this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 
the density functions of the distributions estimated by a distributional forest are plotted together with the 
corresponding observations. This illustrates that appropriate probabilistic forecasts 
are provided by this model as they are close to the observed values. Beside the expected values the predictions 
provided by distributional models offer a wide range of information as they specify the whole distribution and 
therefore enable further methods of inference and analysis. In section \ref{ProbabilisticForecasting} this example 
will be further discussed and it will be shown that the predictions made by the distributional forest are 
reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, 
the scale parameter, and therefore the whole distribution.

@@

%After explaining the novel methodology in Section~\ref{Methodology} of this paper the focus will then be put on an application in the field of probabilistic forecasting in Section~\ref{ProbabilisticForecasting}. The new method is applied on precipitation data in a mountainous region with a very large number of numerical weather prediction quantities and the results are compared to those of other methods commonly used in this research area. In a first setting only one observation station is taken into consideration. The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated by a distributional forest are plotted together with the corresponding observations. This illustrates that well fitting predicted expected values 
%%% well fitting predicted expected values -> appropriate probabilistic forecasts? </TH>
%are provided by this model as they are close to the observed values. Beside the expected values the predictions provided by distributional models offer a wide range of information as they specify the whole distribution and therefore enable further methods of inference and analysis. In section \ref{ProbabilisticForecasting} this example will be further discussed and it will be shown that the predictions made by the distributional forest are reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, the scale parameter, and therefore the whole distribution.
